---
output: html_document
---
<br>
<br>
<br>
<h2>Does Orange Juice or Ascorbic Acid and Dosage Levels Effect How Fast Teeth Grow?</h2>
<br>
<br>
Russ Robbins
<br>
<br>
<br>
**Overview**
<br>
<br>
This analysis assesses whether two variations in the experiment had an effect to the extent that the findings could be expected to be seen in the general population. The results indicate that one variation (dosage level) has an effect.
<br>
<br>
**Subjects**
<br>
<br>
There were six different groups of ten guinea pigs each. 
<br>
<br>
**Measurements**
<br>
<br>
The before and after length of teeth in millimeters.
<br>
<br>
**Variations in the Experiment**
<br>
<br>
Variation 1: Vitamin C provided using two differing delivery mechanisms (orange juice and absorbic acid).

Variation 2: Three different dosage levels (0.5, 1, and 2 milligrams).
<br>
<br>
**Answers**
<br>
<br>
Question 1: Did guinea pigs that received vitamin C via orange juice grow teeth faster than those who received vitamin C in ascorbic acid? 

Answer: It is possible that guinnea pig populations that receive Vitamin C via orange juice or via ascorbic acid vary in their tooth growth rates. However, the Vitamin C delivery mechanism data in this data set do not show a statistically significant difference in the growth rates. 

Question 2: Did guinea pigs that received different dosage levels (0.5, 1, 2 milligrams) grow teeth at different rates?

Answer: It is entirely possible that guinnea pig populations that are provided Vitamin C at 2 mm per dose have faster growing teeth then those given 1 mm per dose. This relationship also holds between guinea pigs with 1 mm vs. 005 mm dose amounts.
<br>
<br>
**Caveats**
<br>
<br>
Note that these results are tentative because of the following reason. The sample sizes, ten in any group, is very small. Larger sample sizes would have provided more information that could be relied upon to make stronger claims. While the possible error related to accepting the null hypotheses incorrectly (that there wasn't a difference between any two groups) was considered, the possible error related to accepting the alternative hypothesis incorrectly, was not considered. It was not considered, because the standard method for addressing this risk, is increasing sample size. In this situation, the data was provided to I the analyst, and I did not have access to the experiment which occurred before 1952, which was fourteen years before I was born. Nonetheless, in an ideal world, I would have rerun the experiment with many more guinnea pigs.
<br>
<br>
**Technical Discussion**
<br>
<br>
This section discusses the results after testing six hypotheses. The results are summarized in the table below.
<br>
<br>

<!-- Start Styles. Move the 'style' tags and everything between them to between the 'head' tags -->
<style type="text/css">
.myTable { width:100%;background-color:#eee;border-collapse:collapse; }
.myTable th { background-color:#000;color:white; }
.myTable td, .myTable th { padding:5px;border:1px solid #000; }
</style>
<!-- End Styles -->
<table class="myTable">
<tr>
<th >#</th><th>Null</th><th>Alt</th><th>t-score</th><th>t-crit</th>
<th>p-value</th><th>alpha</th><th>conf-int</th><th>Reject Null?</th>
</tr>
<tr>
<td>1</td><td>OJ = VC</td><td>OJ not = VC</td>
<td>1.92</td><td>2.00</td>
<td>0.06</td><td>.05</td>
<td>-0.16, 7.56</td><td>no</td>
</tr>
<tr>
<td>2</td><td>OJ <= VC</td><td>OJ > VC</td>
<td>1.92</td><td>1.67</td>
<td>.03</td><td>.05</td>
<td>0.4708, Inf</td><td>yes</td>
</tr>
<tr>
<td>3</td><td>2mm = 1mm</td><td>2mm not = 1mm</td>
<td>4.90</td><td>2.00</td>
<td><.00001</td><td>.05</td>
<td>3.74, 8.99</td><td>yes</td>
</tr>
<tr>
<td>4</td><td>2mm <= 1mm</td><td>2mm > 1mm</td>
<td>4.90</td><td>1.67</td>
<td><.00001</td><td>.05</td>
<td>4.17, Inf</td><td>yes</td>
</tr>
<tr>
<td>5</td><td>1mm = 0.5mm</td><td>1mm not = 0.5mm</td><td>6.48</td><td>2.00</td><td><.00001</td><td>.05</td>
<td>6.28, 11.98</td><td>yes</td>
</tr>
<tr>
<td>6</td><td>1mm <= 0.5mm</td><td>1mm > 0.5mm</td><td>6.48</td><td>1.67</td><td><.000001</td><td>.05</td>
<td>6.75, Inf</td><td>yes</td>
</tr>
</table>
<center><h4>Table 1: Hypotheses Tests Results</h4></center>
--------------------

**Hypothesis 1**: Did guinea pigs that received vitamin C via orange juice grow teeth faster than those who received vitamin C in ascorbic acid? 

No. When the two groups' averages were compared, they were not statistically different, since t-score is not greater than t-critical, the p-value is greater than alpha, and mean difference of zero is within the confidence interval. 

**Hypothesis 2**: Was the mean for the subjects receiving orange juice greater than the mean of those subjects receiving ascorbic acid?

A small effect was seen, as evidenced by the t-score that is somewhat greater than the t-critical, the p-value which is somewhat less than alpha, and the fact that a mean difference of zero was not in the confidence interval. However, the confidence interval of prospective mean differences did reach close to zero. Since the confidence interval almost included zero, we should be careful about indicating that there is an effect, expecially since Hypothesis 1 was found faulty.

**Hypothesis 3**: Did the groups receiving 2mm and 1mm doses have a difference in mean tooth growth?

Yes. The two groups (those receiving 2mm and those receiving 1mm doses) did not have a difference of means that could be explained only by chance. The t-score value was greater than the t-critical value, the p-value was less than the alpha value, and the estimated difference in means ranged from 3.74 mm to 8.99 mm.

**Hypothesis 4**: Did the groups receiving 2mm have greater mean tooth growth than subjects that received 1mm doses?

Yes. The t-score value was greater than the t-critical value, the p-value was less than the alpha value, and the estimated difference in means was greater than 4.17 mm.

**Hypothesis 5**: Did subjects receiving 2mm and those receiving 1mm doses have a statistically significant difference of means?

Yes. The t-score value was greater than the t-critical value, the p-value was less than the alpha value, and the estimated difference in means ranged from 6.28 mm to 11.98 mm.

**Hypothesis 6**: Did the groups receiving 1mm have greater mean tooth growth than subjects that received 0.5mm doses?

Yes, and the test indicated that the difference between 0.5 mm and 1 mm doses may have a larger effect than the doses ranging from 1mm to 2mm.
<br>
<br>
**Code**
<br>
<br>
1. Load the toothgrowth data
```{r}
library(knitr); data(ToothGrowth); attach(ToothGrowth);
unique(supp); unique(dose)
```

2. Provide a basic summary of the data.
```{r}
summary(ToothGrowth)
```
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering). 

```{r}
VC<-subset(ToothGrowth,supp=="VC"); OJ<-subset(ToothGrowth,supp=="OJ")
Zero.5<-subset(ToothGrowth,dose==0.5); One<-subset(ToothGrowth,dose==1.0)
Two<-subset(ToothGrowth,dose==2.0)
meanVC<-mean(VC$len); meanOJ<-mean(OJ$len)
meanZero.5<-mean(Zero.5$len); meanOne<-mean(One$len); meanTwo<-mean(Two$len)
```
Hypothesis 1: &nbsp;&nbsp;|&nbsp; Null: meanOJ - meanVC =  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-OJ - Mean_VC != 0
```{r}
ojEqVc<-t.test(x=OJ$len,y=VC$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("two.sided"))
ojEqVc
```
Hypothesis 2: &nbsp;&nbsp;|&nbsp; Null: meanOJ - meanVC <=  0 &nbsp;&nbsp;|&nbsp; Alternative: meanOJ - meanVC >  0
```{r}
ojGtVc<-t.test(x=OJ$len,y=VC$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("greater"))
ojGtVc
```
Hypothesis 3: &nbsp;&nbsp;|&nbsp; Null: mean2 - mean1 =  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-2 - Mean_1 != 0
```{r}
twoEqOne<-t.test(x=Two$len,y=One$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("two.sided"))
twoEqOne
```
Hypothesis 4: &nbsp;&nbsp;|&nbsp; Null: mean2 - mean1 <=  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-2 - Mean_1 > 0
```{r}
twoGtOne<-t.test(x=Two$len,y=One$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("greater"))
twoGtOne
```
Hypothesis 5:&nbsp;&nbsp;|&nbsp; Null: mean-1 - mean-Zero.5 =  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-1 - Mean-Zero.5 != 0
```{r}
oneEqZero.5<-t.test(x=One$len,y=Zero.5$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("two.sided"))
oneEqZero.5
```
Hypothesis 6: &nbsp;&nbsp;|&nbsp; Null: mean-1 - mean-Zero.5 <=  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-1 - Mean-Zero.5 > 0
```{r}
oneGtZero.5<-t.test(x=One$len,y=Zero.5$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("greater"))
oneGtZero.5
```

