---
output: html_document
---

<h2>Does Orange Juice or Ascorbic Acid and Dosage Levels Effect How Fast Teeth Grow?</h2>
Author: Russ Robbins

<a href='https://github.com/robbinsr/compare_approaches/blob/master/README.md'>Code Repository (right click, open new window or tab)</a>
<br>
<br>
<h4>**Special Note:** Despite my use of this odd data set, this work shows a firm understanding of comparing two approaches and determining whether each had an effect that couldn't happen just by chance. Thanks for bearing with me.</h4>
<br>
<br>
<br>
<hr>
**Overview**
<br>
<br>


This analysis determines whether different approaches for increasing how fast teeth grow actually work. The results indicate that one kind of approach (increasing dosage level) has an effect. In fact increasing a 0.5mm dose to a 1mm does has a greater effect on teeth growing than a dose change from 1mm to 2mm.

+ In real life, a comparable finding could be that we discovered that increasing the number of ads from 500 to 1000 has a greater effect on customers purchasing a streaming service than increasing the number of ads from 1000 to 2000.

**Subjects**
<br>
<br>
There were six different groups of ten guinea pigs each.

+ In real life this could be a sample of particular segment of customers. 

**Measurements**
<br>
<br>
The before and after length of teeth in millimeters. 

+ In real life this could be the number minutes our randomly sampled customers, of a certain demographic, used streaming TV.


**Approaches**
<br>
<br>
Question 1: Compare two delivery mechanisms. Vitamin C provided using two differing delivery mechanisms (orange juice and absorbic acid). 

+ In real life, this could be comparing two marketing approaches; TV vs. Radio ads.

Question 2: Compare three different dosage levels (0.5, 1, and 2 milligrams). 

+ In real life this could be number of ads within a certain time period.

**Answers**

Question 1: Did guinea pigs that received vitamin C via orange juice grow teeth faster than those who received vitamin C in ascorbic acid?

+ In real life: Is there a difference in the amount of time customers spent using our streaming service, when subjects were presented either with TV or radio ads but not both?

Answer: It is possible that guinnea pig populations that receive Vitamin C via orange juice or via ascorbic acid vary in their tooth growth rates. However, the Vitamin C delivery mechanism data in this data set do not show a statistically significant difference in the growth rates.

+ The real life comparable result could be "There is no statistically different difference in consumers usage/purchase behaviors when we use TV and Radio ads, if we control for all other factors.""

Question 2: Did guinea pigs that received different dosage levels (0.5, 1, 2 milligrams) grow teeth at different rates? 

+ In real life: "Did the number of TV ads or Radio ads affect the amount of streaming used?"

Answer: It is entirely possible that guinnea pig populations that are provided Vitamin C at 2 mm per dose have faster growing teeth then those given 1 mm per dose. This relationship also holds between guinea pigs with 1 mm vs. 005 mm dose amounts.

+ The real life analog could be moving from 500 ads to 1000 ads increases has a larger effect on streaming than for the increase in ads from 1000 to 2000.

<br>
<hr>
At this point, for simplicity, I will stop showing how what I did with guinea pigs can also be done with customers.
----
<hr>
<br>
**Caveats**
<br>
<br>
Note that these results are tentative because of the following reason. The sample sizes, ten in any group, is very small. Larger sample sizes would have provided more information that could be relied upon to make stronger claims. While the possible error related to accepting the null hypotheses incorrectly (that there wasn't a difference between any two groups) was considered, the possible error related to accepting the alternative hypothesis incorrectly, was not considered. It was not considered, because the standard method for addressing this risk, is increasing sample size. In this situation, the data was provided to I the analyst, and I did not have access to the experiment which occurred before 1952, which was fourteen years before I was born. Nonetheless, in an ideal world, I would have rerun the experiment with many more guinnea pigs.
<br>
<br>
**Technical Discussion**
<br>
<br>
This section discusses the results after testing six hypotheses. The results are summarized in the table below.
<br>
<br>

<!-- Start Styles. Move the 'style' tags and everything between them to between the 'head' tags -->
<style type="text/css">
.myTable { width:100%;background-color:#eee;border-collapse:collapse; }
.myTable th { background-color:#000;color:white; }
.myTable td, .myTable th { padding:5px;border:1px solid #000; }
</style>
<!-- End Styles -->
<table class="myTable">
<tr>
<th >#</th><th>Null</th><th>Alt</th><th>t-score</th><th>t-crit</th>
<th>p-value</th><th>alpha</th><th>conf-int</th><th>Reject Null?</th>
</tr>
<tr>
<td>1</td><td>OJ = VC</td><td>OJ not = VC</td>
<td>1.92</td><td>2.00</td>
<td>0.06</td><td>.05</td>
<td>-0.16, 7.56</td><td>no</td>
</tr>
<tr>
<td>2</td><td>OJ <= VC</td><td>OJ > VC</td>
<td>1.92</td><td>1.67</td>
<td>.03</td><td>.05</td>
<td>0.4708, Inf</td><td>yes</td>
</tr>
<tr>
<td>3</td><td>2mm = 1mm</td><td>2mm not = 1mm</td>
<td>4.90</td><td>2.00</td>
<td><.00001</td><td>.05</td>
<td>3.74, 8.99</td><td>yes</td>
</tr>
<tr>
<td>4</td><td>2mm <= 1mm</td><td>2mm > 1mm</td>
<td>4.90</td><td>1.67</td>
<td><.00001</td><td>.05</td>
<td>4.17, Inf</td><td>yes</td>
</tr>
<tr>
<td>5</td><td>1mm = 0.5mm</td><td>1mm not = 0.5mm</td><td>6.48</td><td>2.00</td><td><.00001</td><td>.05</td>
<td>6.28, 11.98</td><td>yes</td>
</tr>
<tr>
<td>6</td><td>1mm <= 0.5mm</td><td>1mm > 0.5mm</td><td>6.48</td><td>1.67</td><td><.000001</td><td>.05</td>
<td>6.75, Inf</td><td>yes</td>
</tr>
</table>
<center><h4>Table 1: Hypotheses Tests Results</h4></center>
--------------------

**Hypothesis 1**: Did guinea pigs that received vitamin C via orange juice grow teeth faster than those who received vitamin C in ascorbic acid? 

No. When the two groups' averages were compared, they were not statistically different, since t-score is not greater than t-critical, the p-value is greater than alpha, and mean difference of zero is within the confidence interval. 


**Hypothesis 2**: Was the mean for the subjects receiving orange juice greater than the mean of those subjects receiving ascorbic acid?

A small effect was seen, as evidenced by the t-score that is somewhat greater than the t-critical, the p-value which is somewhat less than alpha, and the fact that a mean difference of zero was not in the confidence interval. However, the confidence interval of prospective mean differences did reach close to zero. Since the confidence interval almost included zero, we should be careful about indicating that there is an effect, expecially since Hypothesis 1 was found faulty.


**Hypothesis 3**: Did the groups receiving 2mm and 1mm doses have a difference in mean tooth growth?

Yes. The two groups (those receiving 2mm and those receiving 1mm doses) did not have a difference of means that could be explained only by chance. The t-score value was greater than the t-critical value, the p-value was less than the alpha value, and the estimated difference in means ranged from 3.74 mm to 8.99 mm.


**Hypothesis 4**: Did the groups receiving 2mm have greater mean tooth growth than subjects that received 1mm doses?

Yes. The t-score value was greater than the t-critical value, the p-value was less than the alpha value, and the estimated difference in means was greater than 4.17 mm.

**Hypothesis 5**: Did subjects receiving 2mm and those receiving 1mm doses have a statistically significant difference of means?

Yes. The t-score value was greater than the t-critical value, the p-value was less than the alpha value, and the estimated difference in means ranged from 6.28 mm to 11.98 mm.


**Hypothesis 6**: Did the groups receiving 1mm have greater mean tooth growth than subjects that received 0.5mm doses?

Yes, and the test indicated that the difference between 0.5 mm and 1 mm doses may have a larger effect than the doses ranging from 1mm to 2mm.
<br>
<br>


**Code**
<br>
<br>


1. Load the toothgrowth data
```{r}
library(knitr); data(ToothGrowth); attach(ToothGrowth);
unique(supp); unique(dose)
```

2. Provide a basic summary of the data.
```{r}
summary(ToothGrowth)
```
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering). 


```{r}
VC<-subset(ToothGrowth,supp=="VC"); OJ<-subset(ToothGrowth,supp=="OJ")
Zero.5<-subset(ToothGrowth,dose==0.5); One<-subset(ToothGrowth,dose==1.0)
Two<-subset(ToothGrowth,dose==2.0)
meanVC<-mean(VC$len); meanOJ<-mean(OJ$len)
meanZero.5<-mean(Zero.5$len); meanOne<-mean(One$len); meanTwo<-mean(Two$len)
```
Hypothesis 1: &nbsp;&nbsp;|&nbsp; Null: meanOJ - meanVC =  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-OJ - Mean_VC != 0

```{r}
ojEqVc<-t.test(x=OJ$len,y=VC$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("two.sided"))
ojEqVc
```
Hypothesis 2: &nbsp;&nbsp;|&nbsp; Null: meanOJ - meanVC <=  0 &nbsp;&nbsp;|&nbsp; Alternative: meanOJ - meanVC >  0

```{r}
ojGtVc<-t.test(x=OJ$len,y=VC$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("greater"))
ojGtVc
```
Hypothesis 3: &nbsp;&nbsp;|&nbsp; Null: mean2 - mean1 =  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-2 - Mean_1 != 0

```{r}
twoEqOne<-t.test(x=Two$len,y=One$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("two.sided"))
twoEqOne
```
Hypothesis 4: &nbsp;&nbsp;|&nbsp; Null: mean2 - mean1 <=  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-2 - Mean_1 > 0

```{r}
twoGtOne<-t.test(x=Two$len,y=One$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("greater"))
twoGtOne
```

Hypothesis 5:&nbsp;&nbsp;|&nbsp; Null: mean-1 - mean-Zero.5 =  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-1 - Mean-Zero.5 != 0

```{r}
oneEqZero.5<-t.test(x=One$len,y=Zero.5$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("two.sided"))
oneEqZero.5

```
Hypothesis 6: &nbsp;&nbsp;|&nbsp; Null: mean-1 - mean-Zero.5 <=  0 &nbsp;&nbsp;|&nbsp; Alternative: Mean-1 - Mean-Zero.5 > 0

```{r}
oneGtZero.5<-t.test(x=One$len,y=Zero.5$len,var.equal=TRUE,conf.level=0.95,paired=FALSE,alternative=c("greater"))
oneGtZero.5
```

